# Rcpp

<script src="./js/hideOutput.js"></script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~")
```


Sometimes R code just isn’t fast enough. You’ve used profiling to figure out where your bottlenecks are, and you’ve done everything you can in R, but your code still isn’t fast enough. I
This magic comes by way of the Rcpp package,
Rcpp makes it very simple to connect C++ to R. While it is possible to write C or Fortran code for use in R, it will be painful by comparison. Rcpp provides a clean, approachable API that lets you write high-performance code, insulated from R’s arcane C API. 

Typical bottlenecks that C++ can address include:

    Loops that can’t be easily vectorised because subsequent iterations depend on previous ones.

    Recursive functions, or problems which involve calling functions millions of times. The overhead of calling a function in C++ is much lower than that in R.

    Problems that require advanced data structures and algorithms that R doesn’t provide. Through the standard template library (STL), C++ has efficient implementations of many important data structures, from ordered maps to double-ended queues.

The aim of this chapter is to discuss only those aspects of C++ and Rcpp that are absolutely necessary to help you eliminate bottlenecks in your code. We won’t spend much time on advanced features like object oriented programming or templates because the focus is on writing small, self-contained functions, not big programs. A working knowledge of C++ is helpful, but not essential.

### Premiers programmes en C++

cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')

# add works like a regular R function
add
add(1, 2, 3)

When you run this code, Rcpp will compile the C++ code and construct an R function that connects to the compiled C++ function. We’re going to use this simple interface to learn how to write C++. 



cppFunction('int one() {
  return 1;
}')


This small function illustrates a number of important differences between R and C++:

    The syntax to create a function looks like the syntax to call a function; you don’t use assignment to create functions as you do in R.

    You must declare the type of output the function returns. This function returns an int (a scalar integer). The classes for the most common types of R vectors are: NumericVector, IntegerVector, CharacterVector, and LogicalVector.

    Scalars and vectors are different. The scalar equivalents of numeric, integer, character, and logical vectors are: double, int, String, and bool.

    You must use an explicit return statement to return a value from a function.

    Every statement is terminated by a ;.


We declare the type of each input in the same way we declare the type of the output. While this makes the code a little more verbose, it also makes it very obvious what type of input the function needs.


One big difference between R and C++ is that the cost of loops is much lower in C++. For example, we could implement the sum function in R using a loop. If you’ve been programming in R a while, you’ll probably have a visceral reaction to this function!

cppFunction('double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total;
}')

The for statement has a different syntax: for(init; check; increment). This loop is initialised by creating a new variable called i with value 0. Before each iteration we check that i < n, and terminate the loop if it’s not. After each iteration, we increment the value of i by one, using the special prefix operator ++ which increases the value of i by 1.
In C++, vector indices start at 0. I’ll say this again because it’s so important: IN C++, VECTOR INDICES START AT 0! This is a very common source of bugs when converting R functions to C++.

C++ provides operators that modify in-place: total += x[i] is equivalent to total = total + x[i]. Similar in-place operators are -=, *=, and /=.

sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}

x <- runif(1e3)
microbenchmark(
  sum(x),
  sumC(x),
  sumR(x)
)

cppFunction('NumericVector pdistC(double x, NumericVector ys) {
  int n = ys.size();
  NumericVector out(n);

  for(int i = 0; i < n; ++i) {
    out[i] = sqrt(pow(ys[i] - x, 2.0));
  }
  return out;
}')

C++ uses pow(), not ^, for exponentiation. In C++, you subset a matrix with (), not [].

Use .nrow() and .ncol() methods to get the dimensions of a matrix.


## Ecrire un programme C++ et l'exécuter avec `Rcpp::sourcecpp`

Your stand-alone C++ file should have extension .cpp, and needs to start with:

#include <Rcpp.h>
using namespace Rcpp;

And for each function that you want available within R, you need to prefix it with:

// [[Rcpp::export]]

To compile the C++ code, use sourceCpp("path/to/file.cpp"). This will create the matching R functions and add them to your current session. Note that these functions can not be saved in a .Rdata file and reloaded in a later session; they must be recreated each time you restart R. For example, running sourceCpp() on the following file implements mean in C++ and then compares it to the built-in mean():

```{r, eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double meanC(NumericVector x) {
  int n = x.size();
  double total = 0;

  for(int i = 0; i < n; ++i) {
    total += x[i];
  }
  return total / n;
}

/*** R
library(microbenchmark)
x <- runif(1e5)
microbenchmark(
  mean(x),
  meanC(x)
)
*/
```

## Attributes dans other classes

You’ve already seen the basic vector classes (IntegerVector, NumericVector, LogicalVector, CharacterVector) and their scalar (int, double, bool, String) and matrix (IntegerMatrix, NumericMatrix, LogicalMatrix, CharacterMatrix) equivalents.

Rcpp also provides classes List and DataFrame, but they are more useful for output than input. This is because lists and data frames can contain arbitrary classes but C++ needs to know their classes in advance. If the list has known structure (e.g., it’s an S3 object), you can extract the components and manually convert them to their C++ equivalents with as(). For example, the object created by lm(), the function that fits a linear model, is a list whose components are always of the same type. 